\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-12-03 flux.tex
Include paragraph on surveillance data.

We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
Influenza A/H1N1 evolved at a rate of 0.55 antigenic units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.  
Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) antigenic units per year.
Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward on antigenic dimension 1 has moved down on antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
Influenza B/Yamagata evolves slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
Very little punctuated evolution is apparent in the evolution of B/Yam.

\subsection*{Competition between influenza clades}

@@

\subsection*{Genetic and antigenic data}

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
However, the majority of isolates were present from 2002 to 2007. 
Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.

@@

Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.

>>>>

We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
Influenza A/H1N1 evolved at a rate of 0.55 antigenic units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.  
Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) antigenic units per year.
Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
Influenza B/Yamagata evolves slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
Very little punctuated evolution is apparent in the evolution of B/Yam.

\subsection*{Competition between influenza lineages}

@@

\subsection*{Genetic, antigenic and surveillance data}

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
However, the majority of isolates were present from 2002 to 2007. 
Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.

@@

Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
As an example, one report states "collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive. Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses. Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses."
In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
